---
title: "Week 9"
format: 
  html:
    embed-resources: true
editor: visual
---

## Homework 9

Two different binary outcomes.  

```{r}
#| message: FALSE
#| warning: false
#| cashe: false
#| echo: false

library(tidyverse)
library(gssr)
```

## Binary Outcomes 1

Does having a college degree affect the age of first birth? We define late first birth as having a first child at age 30 or older. First, let’s create a dataset with binary variables for education and late birth.

```{r}
gss2018 <- gss_get_yr(2018)
firstkid <- gss2018 |>
  select(agekdbrn, educ) |>
  mutate(latebirth = if_else(agekdbrn >= 30 , 1, 0),
         college = if_else(educ >= 12, 1, 0)) |>
  drop_na()
```

Outcome: late first birth(Yes=1 / No=0) 

Predictor: having college degree(Yes=1 / No=0)

```{r}
table(firstkid$college, firstkid$latebirth)
```
P(latebirth|college) = 292/(1128+292) = 0.2056
P(latebirth|noncollege) = 20/(225+20) = 0.0816

### Difference in Probability Regression
Difference in Probability = P(latebirth|college) - P(latebirth|noncollege) = 0.2056 - 0.0816 = 0.1240

#### Model
$$
P(\text{latebirth}=1)_i = \alpha + \beta(\text{college}_i)
$$
#### Estimate in R

```{r}
DP <- glm(latebirth ~ college,
          data = firstkid,
          family = binomial(link = "identity")) #this is for calculating the difference of probability of two groups
summary(DP)
```


$$
P(\text{latebirth}=1) = 0.0816 + 0.1240*\text{college}
$$ 
The estimated effect of college on late birth is 0.1240, meaning that having college degree increases the probability of late first birth by 12.4 percentage points.

#### Confidence Interval

We pick 95% confidence interval. Standard error = 0.02052 Confidence interval = 0.1240 +/- 1.96* 0.02052 = (0.0838, 0.1642). 95% probability that the confidence interval contains true population effect. 

### Risk Ratio Regression Risk ratio 
$$
RR = \frac{P(latebirth=1 | college=1)} {P(latebirth=1|college=0)}
$$ 
RR = 0.2056/0.0816 = 2.52

#### Model
$$
log(P(\text{latebirth}=1)_i) = \alpha + \beta(\text{college}_i)
$$


#### Estimate in R
```{r}
RR <- glm(latebirth ~ college,
          data = firstkid,
          family = binomial(link = "log"))
summary(RR)
```

$$
log(P(\text{latebirth}=1)) = -2.5055 + 0.9239*college
$$ 

#### Risk Ratio

```{r}
exp(0.9239)
```

People with college degree are 2.52 times more likely to have late first birth compared to people with non-college degree.

#### Confidence Intervals

We pick 95% confidence interval. Standard error = 0.02052. Confidence interval = 0.9239 +/- 1.96*0.2205 = (0.4917, 1.3561). Convert to risk ratio interval (1.6351, 3.8810).
95% probability that the confidence interval contains true population effect. 

## Binary Outcomes 2

Do gender affect high school degree? 

```{r}
gendereduc <- gss2018 |>
  select(sex, educ) |>
  mutate(gender = if_else(sex == 1 , 1, 0),
         degree = if_else(educ >= 12, 1, 0)) |>
  drop_na()
```

Outcome: having high school degree(Yes=1 / No=0) 

Predictor:gender (Men=1 / Women=0)

```{r}
table(gendereduc$gender, gendereduc$degree)
```
P(degree|men) = 908/(143+908) = 0.8639
P(degree|women) = 1135/(159+1135) = 0.8771

### Difference in Probability Regression
Difference in Probability = P(degree|men) - P(degree|women) = -0.0132

#### Model
$$
P(\text{college}=1)_i = \alpha + \beta(\text{gender}_i)
$$

#### Estimate in R
```{r}
DP2 <- glm(degree ~ gender,
          data = gendereduc,
          family = binomial(link = "identity")) 
summary(DP2)
```

$$
P(\text{college}=1) = 0.8771 - 0.0132*\text{gender}
$$ 
The estimated effect of gender on high school degree is -0.01319, meaning that being a man decreases the probability of having a high school degree by 1.32 percentage points.

#### Confidence Interval

We pick 95% confidence interval. Standard error = 0.02052 Confidence interval = -0.0132 +/- 1.96* 0.0140 = (-0.0406, 0.0142). 95% probability that the confidence interval contains true population effect. Since this confidence interval contains 0, the true effect could be zero. 

### Risk Ratio Regression 

RR = 0.8639/0.8771 = 0.9850

$$
log(P(\text{degree}=1)_i) = \alpha + \beta(\text{gender}_i)
$$

```{r}
# RR2 <- glm(degree ~ gender,
          # data = gendereduc,
          # family = binomial(link = "log"))
# summary(RR2)
# I run into error with this code, suggesting that the probability might be zero.
```


## Binary Outcomes 3
We change education level to college and recalculate

Do gender affect college degree? 

```{r}
collegeeduc <- gss2018 |>
  select(sex, educ) |>
  mutate(gender = if_else(sex == 1 , 1, 0),
         college = if_else(educ >= 16, 1, 0)) |>
  drop_na()
```

Outcome: having college degree(Yes=1 / No=0) 

Predictor:gender (Men=1 / Women=0)

```{r}
table(collegeeduc$gender, collegeeduc$college)
```
P(college|men) = 336/(715+336) = 0.3197
P(college|women) = 427/(867+427) = 0.3300

### Difference in Probability Regression
Difference in Probability = P(college|men) - P(college|women) = -0.0103

#### Model
$$
P(\text{college}=1)_i = \alpha + \beta(\text{gender}_i)
$$

#### Estimate in R
```{r}
DP3 <- glm(college ~ gender,
          data = collegeeduc,
          family = binomial(link = "identity")) 
summary(DP3)
```

$$
P(\text{college}=1) = 0.3300 - 0.0103*\text{gender}
$$ 
The estimated effect of gender on college degree is -0.0103, meaning that being a man decreases the probability of having a college degree by 1.03 percentage points.

#### Confidence Interval

We pick 95% confidence interval. Standard error = 0.0194 Confidence interval = -0.0103 +/- 1.96* 0.0194 = (-0.0483, 0.0277). 95% probability that the confidence interval contains true population effect. Since this confidence interval contains 0, the true effect could be zero. 

### Risk Ratio Regression

RR = 0.3197/0.3300 = 0.9688

#### Model
$$
log(P(\text{college}=1)_i) = \alpha + \beta(\text{gender}_i)
$$

#### Estimate in R
```{r}
RR3 <- glm(college ~ gender,
          data = collegeeduc,
          family = binomial(link = "log"))
summary(RR3)
```

$$
log(P(\text{college}=1)_i) = -1.1087 - 0.0317(\text{gender}_i)
$$

#### Risk Ratio

```{r}
exp(-0.0317)
```

Men are 0.97 times likely to have college degree compared to women.

#### Confidence Intervals

We pick 95% confidence interval. Standard error = 0.0600. Confidence interval = -0.03168 +/- 1.96*0.05995 = (-0.1492, 0.0858). Convert to risk ratio interval (0.8614, 1.0896).
95% probability that the confidence interval contains true population effect; 95% the true risk ratio sets with (0.8614, 1.0896).
