---
title: "Week 9"
format: 
  html:
    embed-resources: true
editor: visual
---

## Homework 9

Two different binary outcomes.  

```{r}
#| message: FALSE
#| warning: false
#| cashe: false
#| echo: false

library(tidyverse)
library(gssr)
```

## Binary Outcomes 1

Does having a college degree affect late first birth age? We define late first birth as people have their first kid at age 30 or above 30. First, Let's create a dataset with binary variables of education and late birth.

```{r}
gss2018 <- gss_get_yr(2018)
firstkid <- gss2018 |>
  select(agekdbrn, educ) |>
  mutate(latebirth = if_else(agekdbrn >= 30 , 1, 0),
         college = if_else(educ >= 12, 1, 0)) |>
  drop_na()
```

Outcome: late first birth(Yes=1 / No=0) Predictor: having college degree(Yes=1 / No=0)

```{r}
table(firstkid$college, firstkid$latebirth)
```

### Difference in Probability Regression

$$
P(\text{latebirth}=1)_i = \alpha + \beta(\text{college}_i)
$$

```{r}
DP <- glm(latebirth ~ college,
          data = firstkid,
          family = binomial(link = "identity")) #this is for calculating the difference of probability of two groups
summary(DP)
```

$$
P(\text{latebirth}=1) = 0.0816 + 0.1240*\text{college}
$$ 
The estimated effect of college on late birth is 0.124, meaning that having college degree increases the probability of late first birth by 12.4 percentage points.

#### Confidence Interval

We pick 95% confidence interval. Standard error = 0.02052 Confidence interval = 0.1240 +/- 1.96\* 0.02052 = (0.0838, 0.1642). 95% probability that the confidence interval contains true population effect. 

### Risk Ratio Regression Risk ratio 
$$
RR = \frac{P(latebirth=1 | college=1)} {P(latebirth=1|college=0)}
$$ 
$$
log(P(\text{latebirth}=1)_i) = \alpha + \beta(\text{college}_i)
$$

```{r}
RR <- glm(latebirth ~ college,
          data = firstkid,
          family = binomial(link = "log"))
summary(RR)
```

$$
log(P(\text{latebirth}=1)) = -2.5055 + 0.9239*college
$$ Risk Ratio

```{r}
exp(0.9239)
```

People with college degree are 2.5 times more likely to have late first birth compared to people with non-college degree.

#### Confidence Intervals

We pick 95% confidence interval. Standard error = 0.02052. Confidence interval = 0.9239 +/- 1.96*0.2205 = (0.4917, 1.3561).
95% probability that the confidence interval contains true population effect. 

## Binary Outcomes 2

Do gender affect college degree? 

```{r}
gendereduc <- gss2018 |>
  select(sex, educ) |>
  mutate(gender = if_else(sex == 1 , 1, 0),
         college = if_else(educ >= 12, 1, 0)) |>
  drop_na()
```

Outcome: having college degree(Yes=1 / No=0) Predictor:gender (Men=1 / Women=0)

```{r}
table(gendereduc$gender, gendereduc$college)
```

### Difference in Probability Regression

$$
P(\text{college}=1)_i = \alpha + \beta(\text{gender}_i)
$$

```{r}
DP2 <- glm(college ~ gender,
          data = gendereduc,
          family = binomial(link = "identity")) 
summary(DP2)
```

$$
P(\text{college}=1) = 0.8771 - 0.0132*\text{gender}
$$ 
The estimated effect of gender on college degree is -0.01319, meaning that being a man decreases the probability of having a college degree by 1.32 percentage points.

#### Confidence Interval

We pick 95% confidence interval. Standard error = 0.02052 Confidence interval = -0.0132 +/- 1.96* 0.0140 = (-0.0406, 0.0142). 95% probability that the confidence interval contains true population effect. Since this confidence interval contains 0, the true effect could be zero. 

### Risk Ratio Regression Risk ratio 

$$
log(P(\text{college}=1)_i) = \alpha + \beta(\text{gender}_i)
$$

```{r}
# RR2 <- glm(college ~ gender,
          # data = gendereduc,
          # family = binomial(link = "log"))
# summary(RR2)
# I run into error with this code, suggesting that the probability might be zero.
```








